Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-04-12 doc/thesis/intro.tex
More writing.

\subsection{Tree shapes}
\label{subsubsec:treeshape}

@@

Phylodynamic methods have been used to investigate epidemiological parameters
such as transmission rate, recovery rate, and basic reproductive
number~\autocite{pybus2009evolutionary, volz2013viral}. These studies make
inferences about epidemiological processes from the genetic diversity of virus
populations, which is usually represented in the form of a phylogeny. The
majority of these employ a Bayesian \gls{MCMC} approach to infer parameters of
an epidemiological model whose likelihood can be calculated, most often some
variation of the birth-death~\autocite{kendall1948generalized} or
coalescent~\autocite{kingman1982coalescent} models.
\textcite{stadler2011estimating} develop a formula for the likelihood of a
phylogeny with heterochronous tips under the birth-death model, which has been
used to estimate the basic reproductive number of several viral
epidemics~\autocite{stadler2011estimating}. However, the birth-death model is
cannot tell us anything about population structure, as it assumes that every
individual becomes infected at the same rate. \textcite{volz2012complex} writes
down the likelihood of a heterochronous phylogeny under a coalescent model with
arbitrarily complex population dynamics. This opens the door to more complex
inferences about population structure, as the population can be partitioned
into compartments with different transmission and recovery rates, but still
assumes that each compartment is homogeneously mixed. In other words, the
coalescent model can tell us about the \emph{global} structure of a population,
such as whether there exists a high-risk subgroup, but not about the
\emph{local} structure, such as the average number of contacts each individual
has.

@@

\section{Approximate Bayesian computation}
\label{sec:abc}
\glsreset{ABC}

@@

To illustrate these concepts, consider the well-known linear model. For
clarity, we will restrict our attention to the case of one-dimensional data and
outcomes where each $x_i$ and $y_i$ is a real number. The linear model
postulates that the outcomes are linearly related to the data, modulo some
noise introduced by measurement error, environmental fluctuations, and other
external factors. Formally, $y_i = \beta x_i + \varepsilon_i$, where $\beta$ is
the slope of the linear relationship, and $\varepsilon_i$ is the error
associated with measurement $i$. We can make this model a statistical one by
hypothesizing a distribution for the error terms $\varepsilon_i$; most
commonly, it is assumed that they are normally distributed with variance
$\sigma$. In mathematical terms, $Y_i \sim \beta x_i + \N(0, \sigma^2)$, where
``$\sim$'' means ``is distributed as''. We can see from this formulation that
the model is parametric, with parameters $\beta$ and $\sigma$. Moreover, we can
write down the probability density of observing outcome $y_i$ given the
parameters,
\[
  \Pr(Y_i = y_i \mid \beta, \sigma) = 
  \Pr\left(\N(0, \sigma^2) = y_i - \beta x_i\right).
\]
Note that we have followed the standard statistical abuse of notation and used
``$\Pr$'' to refer to a probability \emph{density}, rather than a probability.
Assuming all the $y_i$ are independent, the probability density of the entire
observed set of outcomes is the product of the probability density of each
individual $y_i$,
\[
  \Pr(\vec{Y} = \vec{y} \mid \beta, \sigma) = 
  \prod_{i=1}^N \Pr(Y_i = y_i \mid \beta, \sigma).
\]

For a general model, the probability density of $\vec{y}$ given the parameters
$\theta$ is also known as the \defn{likelihood}, written $\L$, of $\theta$.
That is, $\L(\theta \mid \vec{y}) = \Pr(\vec{y} \mid \theta)$. The higher the
value of the likelihood, the more likely the observations $\vec{y}$ are under
the model. Thus, the likelihood provides a natural criterion for fitting the
model parameters: we want to pick $\theta$ such that the probability density of
our observed outcomes $\vec{y}$ is as high as possible. The parameters which
optimize the likelihood are known as the \textit{\gls{ML}} estimates, denoted
$\hat{\theta}$. \Gls{ML} estimation is usually performed with numerical
optimization. In the simplest terms, many possible values for $\theta$ are
examined, $\L(\theta \mid \vec{y})$ is calculated for each, and the parameters
which produce the highest value are accepted. Many sophisticated numerical
optimization methods exist, although they may not be guaranteed to find the
true \gls{ML} estimates if the likelihood function is complex. Occasionally, as
in the case of least squares, the \gls{ML} estimates can be found explicitly by
setting the likelihood function's derivatives to zero.

\Gls{ML} estimation makes use only of the data and outcomes to estimate the
model parameters $\theta$. However, it is frequently the case that the
investigator has some additional information or \defn{prior belief} about what
$\theta$ are likely to be. For example, in the linear regression case, the
instrument used to measure the outcomes may have a well-known margin of error,
or the sign of the slope may be obvious from previous experiments. The Bayesian
approach to model fitting makes use of this information by codifying the
investigator's beliefs as a \defn{prior distribution} on the parameters,
denoted $\Pr(\theta)$. Instead of considering only the likelihood, Bayesian
inference focuses on the product of the likelihood and the prior, $\Pr(\vec{y}
\mid \theta) \Pr(\theta)$. Bayes' theorem tells us that this product is related
to the \textit{posterior distribution} on $\theta$,
\[
  \Pr(\theta \mid \vec{y}) = \frac{\Pr(\vec{y} \mid \theta) \Pr(\theta)}
                                  {\Pr(\vec{y})}.
\]

@@

A popular method for circumventing the normalizing constant is the use of
\gls{MCMC} to obtain a sample from the posterior distribution. \Gls{MCMC} works
by defining a Markov chain whose states are indexed by possible model
parameters. The transition probability from state $\theta_1$ to state
$\theta_2$ is taken to be
\[
  \max\left(1, \frac{\Pr(\vec{y} \mid \theta_2) \Pr(\theta_2) q(\theta_2, \theta_1)}
                    {\Pr(\vec{y} \mid \theta_1) \Pr(\theta_2) q(\theta_1, \theta_2)} \right),
\]
where $q(\theta, \theta')$ is a symmetric \defn{proposal distribution} used in
the algorithm to generate the chain. The stationary distribution of this Markov
chain is equal to the posterior distribution on $\theta$. Therefore, if a long
enough random walk is performed on the chain, the distribution of states
visited will be a Monte Carlo approximation of $\Pr(\theta \mid \vec{y})$, from
which we can calculate statistics of interest. Actually performing this random
walk is straightforward and is accomplished via the Metropolis-Hastings
algorithm (\cref{alg:mh}).

\begin{algorithm}
  \label{alg:mh}
  \caption{Metropolis-Hastings algorithm for Markov chain Monte Carlo.}
  \begin{algorithmic}
    \State Draw $\theta$ according to $\Pr(\theta)$
    \Loop
      \State Propose $\theta'$ according to $q(\theta, \theta')$
      \State Accept $\theta \gets \theta'$ with probability
      $\max \left( 1, 
       \dfrac{\Pr(\vec{y} \mid \theta') \Pr(\theta') q(\theta', \theta)}
             {\Pr(\vec{y} \mid \theta\phantom{'}) \Pr(\theta\phantom{'}) q(\theta, \theta')}
       \right)$
    \EndLoop
  \end{algorithmic}
  \label{alg:mh}
\end{algorithm}

\subsection{Overview and motivation for ABC}
\label{subsec:abcoverview}

Most mathematical models are amenable to fitting via one or both of the
approaches, \gls{ML} or Bayesian inference, discussed above. However, there are
some, particularly in the domain of population genetics, for which calculation
of either the likelihood or the product of the likelihood and the prior may be
infeasible. For example, one or both of these quantities may be expressible
only as an intractable integral. \Gls{ABC} is designed for such cases, where
standard likelihood-based techniques for model fitting cannot be applied. Such
models are particularly prevalent in population
genetics~\autocite{beaumont2002approximate, beaumont2010approximate}. We will
begin this subsection by describing \gls{ABC} in general terms. Then, we will
use linear regression as a toy problem to demonstrate how \gls{ABC} could be
applied. Finally, we will give an example of a model which is impractical to
fit using likelihood-based methods, and show how its parameters were estimated
using \gls{ABC}.

Ordinarily, Bayesian inference targets the posterior distribution.

%traditional likelihood criterion
%for model ``goodness'' with one based on simulated data generated by the model.
%Good models should be able to simulate data which closely resemble reality. In
%\gls{ABC}, we are no longer interested in the posterior distribution, but
%rather in the distribution of model parameters which produce data sufficiently
%``close'' to the real data. To formalize this, let $d(\cdot, \cdot)$ be a
%distance function on datasets, and $\varepsilon$ be a tolerance level.

Of course, linear models can be fit more easily using likelihood-based methods, 
since the likelihood $\L(\theta \mid \vec{y})$ can be easily calculated (see
\cref{subsec:fit}). One model where this is not the case is Kingman's
coalescent~\autocite{kingman1982coalescent}.

As an example of such a model, we will review the first practical application
of \gls{ABC} by \textcite{tavare1997inferring}. The aim of the investigation
was to calculate the posterior distribution on the coalescence time $t$ of some
observed sequences $D$ under the coalescent model
\autocite{kingman1982coalescent} with the infinite-sites assumption
\autocite{watterson1975number}. Their estimation method relied on two key
observations. First, the number $k$ of \defn{segregating sites} in the data is
informative about the coalescence time while being much easier to work with
than the data itself. A segregating site is a position which is polymorphic in
the observed sequences. Using this observation, the authors replaced their
posterior distribution of interest, $\Pr(t \mid D)$, by
\[
  \Pr(t \mid k) \propto \Pr(t) \Pr(k \mid t),
\]
where again we abuse $\Pr$ to refer to a probability density. Second,
the authors observed that the expected number of segregating sites depends on
the total branch length $l$ of the tree, rather than its height. This enabled
them to write an alternative expression for the product of the likelihood and
prior in terms of $l$,
\[
  \Pr(t) \Pr(k \mid t) = \int_0^\infty \Pr(t, l) \Pr(k \mid l) \d l.
\]
While this integral may be difficult to evaluate analytically, the power of
this formulation is the fact that it is straightforward to \emph{simulate}
values of $t$ and $l$ according to $\Pr(t, l)$, their joint probability under
the coalescent model. One simply runs the (stochastic) coalescent process and
observing the values $\hat{t}$ and $\hat{l}$ which result. It is
straightforward to calculate $\Pr(k \mid \hat{l})$, the probability
that $k$ segregating sites were observed in a tree with total branch length
$\hat{l}$, using the Poisson distribution. Hence if $\hat{l}$ and $\hat{t}$,
having been generated according to $\Pr(t, l)$, are sampled with probability
proportional to $\Pr(k \mid \hat{l})$, they represent an unbiased sample from
the above integral. Taking a large number of such samples allowed the authors
to obtain a Monte Carlo estimate of $\Pr(t \mid k)$.

%\Gls{ABC} is a statistical method designed to fit complex models, which cannot
%be fit using more conventional methods, to observed
%data~\autocite{marin2012approximate, sunnaker2013approximate,
%beaumont2010approximate}. We shall make this precise below, but to fully
%describe and motivate \gls{ABC}, it is necessary to first explain exactly what
%we mean by ``fitting'' a model. We will illustrate this concept with one of the
%simplest and most ubiquitous models: linear regression.
%
%An approximate Bayesian computation approach to least-squares regression might
%be as follows. We will need to make an additional assumption: that the errors
%$\varepsilon_i$ are normally distributed. That is, our linear model is now of
%the form
%\[
%  y_i = bx_i + \N(0, \sigma),
%\]
%where $\N(0, \sigma)$ indicates a normal distribution with mean zero and
%variance $\sigma$. We do not know $\sigma$ in advance, so this is another
%parameter we will need to estimate. There is nothing special about normal
%distributions - we could have chosen any other distribution we deemed
%appropriate. 
%

\subsection{Algorithms for ABC}
\label{subsubsec:abcalg}

\Gls{ABC} does not refer to a particular procedure for model fitting. Rather,
\gls{ABC} refers to the general strategy of choosing model parameters based on
the resulting model's propensity to generate data resembling the real data.
There are three main classes of \gls{ABC} algorithm which have been developed
so far: rejection, \gls{MCMC}, and \gls{SMC}.

All of these approaches require some common elements. First, as with all
Bayesian methods, we are required to specify a \defn{prior distribution},
denoted $\pi$, on the parameter space. The prior specifies what we already know
or believe about the model parameters. Second, in order to compare simulated to
observed data, we need to be able to summarize a data set in a numerical
format. This is accomplished by a function, denoted $\eta$, which computes a
vector of hopefully informative summary statistics on a data set. Third, we
need a distance function $\rho$ which tells us how similar two data sets are to
each other, based on their summary statistics.

>>>>

\subsection{Tree shapes}
\label{subsec:treeshape}

@@

Phylodynamic methods have been used to investigate epidemiological parameters
such as transmission rate, recovery rate, and basic reproductive
number~\autocite{pybus2009evolutionary, volz2013viral}. These studies make
inferences about epidemiological processes from the genetic diversity of virus
populations, which is usually represented in the form of a phylogeny. The
majority of these employ a Bayesian \gls{MCMC} approach to infer parameters of
an epidemiological model whose likelihood can be calculated, most often some
variation of the birth-death~\autocite{kendall1948generalized} or
coalescent~\autocite{kingman1982coalescent} models.
\textcite{stadler2012estimating} develop a formula for the likelihood of a
phylogeny with heterochronous tips under the birth-death model, which has been
used to estimate the basic reproductive number of several viral
epidemics~\autocite{stadler2012estimating}. However, the birth-death model is
cannot tell us anything about population structure, as it assumes that every
individual becomes infected at the same rate. \textcite{volz2012complex} writes
down the likelihood of a heterochronous phylogeny under a coalescent model with
arbitrarily complex population dynamics. This opens the door to more complex
inferences about population structure, as the population can be partitioned
into compartments with different transmission and recovery rates, but still
assumes that each compartment is homogeneously mixed. In other words, the
coalescent model can tell us about the \emph{global} structure of a population,
such as whether there exists a high-risk subgroup, but not about the
\emph{local} structure, such as the average number of contacts each individual
has.

@@

\section{Sequential Monte Carlo}
\label{sec:smc}

\Gls{SMC} is a statistical inference method which samples from a sequence of
probability distributions in a fixed order~\autocite{del2006sequential}, based
on a technique called \gls{SIS}.

\subsection{Sequential importance sampling}
\label{subsec:sis}

The basis of \gls{SIS} is \gls{IS}, which is a method of estimating summary
statistics of distributions which are known only up to a normalizing constant,
and therefore cannot be sampled from directly. That is, if $\pi$ is such a
distribution and $f$ is any real-valued function, \gls{IS} is concerned with
estimating
\[
  \pi(f) = \int f(x)\pi(x)\d x = \int f(x) \frac{\gamma(x)}{Z} \d x,
\]
where the integral is over the space on which $\pi$ is defined, $\gamma(x)$ is
known pointwise, and $Z = \int \gamma(x) \d x$ is the unknown normalizing
constant. Suppose we have at hand another distribution $\eta$, called the
\defn{importance distribution}, from which we are able to sample. Define the
\defn{importance weight} as the ratio ratio $w(x) = \gamma(x)/\eta(x)$. We can
express the normalizing constant $Z$ in terms of the importance weight and
distribution, $Z = \int w(x) \eta(x) \d x$, and in turn write the expectation
of interest as
\[
  \int f(x) \pi(x) \d x = \frac{\int f(x) \gamma(x) \d x}
                               {\int w(x) \eta(x) \d x}.
\]
If we sample a large number of points from $\eta$, then $\eta(x)$ can be
approximated by a Monte Carlo estimate. Since the remaining quantities $f$,
$\gamma$, and $w$ can all be evaluated pointwise, these are all the ingredients
we need to obtain an estimate of $\pi(f)$. Although this is a simple and
elegant approach, the drawback is that the variance of the estimate is
proportional to the variance of the importance weights~\autocite{liu2008monte},
which may be quite large if $\eta$ and $\gamma$ are very different. Therefore,
the practical use of \gls{IS} on its own is limited, since it depends on
finding an importance distribution similar to $\pi$, which we usually know very
little about \textit{a priori}.

\gls{SIS} applies \gls{IS} to the problem of sampling from a distribution $\pi$ 
on an high-dimensional space, say $\pi(\vec{x}) = \pi(x_1, \ldots, x_d)$. The
objective is to build up an importance distribution $\eta$ for $\pi$
sequentially. Let $\vec{x_i} = (x_1, \ldots, x_i)$, for $1 \leq i \leq d$. By
the general product rule,
\[
  \pi(\vec{x}) 
  = \pi(x_1) \pi(x_2 \mid x_1) \cdots
    \pi(x_{d-1} \mid \vec{x_{d-2}}) \pi(x_d \mid \vec{x_{d-1}}).
\]
This decomposition is natural in many contexts, particularly for on-line
estimation. For example, in a stateful model like an \gls{HMM}, $x_i$ may
represent the state at time $i$, with $\pi(\vec{x})$ being the posterior
distribution over possible paths. The importance distribution $\eta$ for $\pi$
will be constructed using a similar decomposition,
\[
  \eta(\vec{x}) 
  = \eta(x_1) \eta(x_2 \mid x_1) \cdots
    \eta(x_{d-1} \mid \vec{x_{d-2}}) \eta(x_d \mid \vec{x_{d-1}}).
\]
The importance weights for $\eta$ with respect to $\pi$ are then defined
recursively as
\begin{align}
  \label{eq:sisw}
  w(\vec{x_i}) = \frac{\pi(\vec{x_i})}{\eta(\vec{x_i})}
  = \frac{\pi(x_i \mid \vec{x_{i-1}})\pi(\vec{x_{i-1}})}
         {\eta(x_i \mid \vec{x_{i-1}})\eta(\vec{x_{i-1}})}
  = \frac{\pi(x_i \mid \vec{x_{i-1}})}
         {\eta(x_i \mid \vec{x_{i-1}})}\cdot w(\vec{x_{i-1}}).
\end{align}
Thus, we can choose $\eta(x_i \mid \vec{x_{i-1}})$ such that the variance of
the importance weights is as small as possible at every step, eventually
arriving at a full importance distribution. The process for choosing $\eta$
will differ according to the problem at hand. One potential choice for
$\eta(x_i \mid \vec{x_{i-1}})$ is simply $\pi(x_i \mid \vec{x_{i-1}})$, if it
is possible to compute. For example, in the case of an \gls{HMM}, $\pi(x_i \mid
\vec{x_{i-1}})$ is simply the transition probability from $x_{i-1}$ to $x_i$.
In a Bayesian setting, $\eta$ may be taken to be the prior distribution. In
some applications, it is possible to compute or approximate the form of $\eta$
which minimizes the variance of the weights. The resulting importance
distribution is known as the \defn{optimal
kernel}~\autocite{cappe2007overview}. Choosing each element of the
decomposition of $\eta$ is a problem-specific task (see \textit{e.g.}
\autocite{smith2013sequential,liu2008monte} for many examples).

The recursive definition \cref{eq:sisw} suggests an algorithm for obtaining a
sample from $\pi$. We begin with a sample of $n$ ``particles'' which have been
sampled from the importance distribution $\eta(x_0)$ for $\pi(x_0)$. The
particles are updated and reweighted $d$ times, corresponding to the $d$
elements of the decomposition of $\pi$. At the $i$th step, each particle is
extended to include $x_i$ drawn from the chosen $\eta(x_i \mid \vec{x_{i-1}})$,
and the importance weights are recalculated and normalized.

Of course, $\eta$ is merely an approximation to $\pi$, and may be a fairly poor
one depending on the application. Try as we might to keep the variances of the
weights low, the cumulative errors at each sequential step tend to push many of
the weights to very low values. This results in a poor approximation to $\pi$,
since only a few particles retain high importance weights after all $d$
sequential steps. To mitigate this problem, we periodically apply a resampling
step when the variance in the importance weights becomes too high. Several
different criteria have been proposed for when to resample, but we focus here
on the one described by \textcite{liu2008monte}, namely the decay of the
\gls{ESS} below a prescribed threshold. The \gls{ESS} of the population of
particles is defined as
\[
  \ESS(w) = \frac{n}{1 + \Var(w)},
\]
where $n$ is the number of particles~\autocite{liu2008monte}. When the
\gls{ESS} drops below the threshold, we resample the particles according to
their weights. This results in the removal of low-weight particles from the
population, and also equalizes all the weights. Various resampling strategies
beyond the basic sampling with replacement have been proposed, but we will not
discuss those here. The full \gls{SIS} algorithm with resampling is given as
\cref{alg:sis}. We use a superscript $^{(k)}$ to refer to the $k$th particle,
and for brevity write $w$ for $w(\vec{x_i})$. The variable $T$ is the threshold
for the \gls{ESS}, below which the population is resampled. It is conventional
to take $T = n/2$.

\begin{algorithm}
  \caption{Sequential importance sampling.}
  \begin{algorithmic}
    \For {$k = 1$ to $n$}
      \State Sample $x_0^{(k)}$ from $\eta(x_0)$
      \Comment{Initialize the $k$th particle}
      \State $w^{(k)} \gets \dfrac{\pi\left(x_0^{(k)}\right)}{\eta\left(x_0^{(k)}\right)}$
    \EndFor
    \For {$i = 1$ to $d$}
      \For {$k = 1$ to $n$}
        \State Sample $x_i^{(k)}$ from $\eta\left(x_i \mid \vec{x_{i-1}^{(k)}}\right)$
        \Comment Extend the $k$th particle
        \State $w^{(k)} \gets \dfrac{\pi\left(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}\right)}{\eta\left(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}\right)} \cdot w^{(k)}$
      \EndFor
      \State Normalize the weights so that $\sum w = 1$
      \If {$\ESS(w) < T$}
        \State Resample the particles according to $w$
        \For {$k = 1$ to $n$}
          \State $w^{(k)} \gets 1/n$
        \EndFor
      \EndIf
    \EndFor
    \State Sample $n$ particles with probabilities $w$
  \end{algorithmic}
  \label{alg:sis}
\end{algorithm}

\subsection{Sequential Monte Carlo samplers}

The \gls{SIS} algorithm described above aims to sample from a high-dimensional
distribution $\pi(x)$, by sequentially sampling from $d$ distributions of lower
but increasing dimension. \textcite{del2006sequential} developed an
\defn{\gls{SMC} sampler} with an alternative objective: to sample sequentially
from $d$ distributions $\pi_1, \ldots, \pi_d$, all of the same dimension and
defined on the same space. The $\pi_i$ are assumed to form a related sequence,
such as posterior distributions attained by sequentially considering new
evidence. As with \gls{SIS}, we assume that $\pi_i = \gamma_i / Z_i$, where
$\gamma_i$ is known pointwise and the normalizing constant $Z_i$ is unknown.

Both algorithms follow the same general pattern: the target distribution is
attained by progressing through a sequence of intermediate distributions. For
\gls{SIS}, the intermediate distributions are lower-dimensional marginals of
the target, while for the \gls{SMC} sampler, they are of the same dimension and
constitute a smooth progression from an initial distribution to the target. In
both cases, the neighbouring distributions in the sequence are related to each
other in some way, and we can take advantage of that relationship to create a
sequence of importance distributions alongside the sequence of targets. In
\gls{SIS}, the neighbouring marginals $\pi(\vec{x_i})$ and
$\pi(\vec{x_{i+1}})$ were related by the conditional density $\pi(x_i \mid
\vec{x_{i-1}})$, which we used to inform the importance distribution.

The sequence of targets considered by the \gls{SMC} sampler are assumed to be
closely enough related, that we can shift an importance distribution for
$\pi_i$ into one for $\pi_{i+1}$ by perturbing the particles in a systematic
way.

\section{Approximate Bayesian computation}
\label{sec:abc}

\glsreset{ABC}

@@

To illustrate these concepts, consider the well-known linear model. For
clarity, we will restrict our attention to the case of one-dimensional data and
outcomes where each $x_i$ and $y_i$ is a single real number. The linear model
postulates that the outcomes are linearly related to the data, modulo some
noise introduced by measurement error, environmental fluctuations, and other
external factors. Formally, $y_i = \beta x_i + \varepsilon_i$, where $\beta$ is
the slope of the linear relationship, and $\varepsilon_i$ is the error
associated with measurement $i$. We can make this model a statistical one by
hypothesizing a distribution for the error terms $\varepsilon_i$; most
commonly, it is assumed that they are normally distributed with variance
$\sigma$. In mathematical terms, $Y_i \sim \beta x_i + \N(0, \sigma^2)$, where
``$\sim$'' means ``is distributed as''. We can see from this formulation that
the model is parametric, with parameters $\theta$ = ($\beta$, $\sigma$).
Moreover, we can write down the probability density of observing outcome $y_i$
given the parameters,
\[
  \Pr(Y_i = y_i \mid \beta, \sigma) = 
  \Pr\left(\N(0, \sigma^2) = y_i - \beta x_i\right).
\]
Note that we have followed the standard statistical abuse of notation and used
``$\Pr$'' to refer to a probability \emph{density}, rather than a probability.
Also, we are treating the $x_i$ as fixed quantities, and therefore have not
conditioned the probability density on $\vec{x}$. Assuming all the $y_i$ are
independent, the probability density of the entire observed set of outcomes is
the product of the probability density of each individual $y_i$,
\[
  \Pr(\vec{Y} = \vec{y} \mid \beta, \sigma) = 
  \prod_{i=1}^N \Pr(Y_i = y_i \mid \beta, \sigma).
\]

For a general model, the probability density of $\vec{y}$ given the parameters
$\theta$ is also known as the \defn{likelihood}, written $\L$, of $\theta$.
That is, $\L(\theta \mid \vec{y}) = \Pr(\vec{y} \mid \theta)$. The higher the
value of the likelihood, the more likely the observations $\vec{y}$ are under
the model. Thus, the likelihood provides a natural criterion for fitting the
model parameters: we want to pick $\theta$ such that the probability density of
our observed outcomes $\vec{y}$ is as high as possible. The parameters which
optimize the likelihood are known as the \textit{\gls{ML}} estimates, denoted
$\hat{\theta}$. That is,
\[
  \hat{\theta} = \argmax_\theta\; \Pr(\vec{y} \mid \theta).
\]
\Gls{ML} estimation is usually performed with numerical optimization. In the
simplest terms, many possible values for $\theta$ are examined, $\L(\theta \mid
\vec{y})$ is calculated for each, and the parameters which produce the highest
value are accepted. Many sophisticated numerical optimization methods exist,
although they may not be guaranteed to find the true \gls{ML} estimates if the
likelihood function is complex. Occasionally, as is the case with linear
regression, the \gls{ML} estimates can be found explicitly by setting the
likelihood function's derivatives to zero.

\Gls{ML} estimation makes use only of the data and outcomes to estimate the
model parameters $\theta$. However, it is frequently the case that the
investigator has some additional information or belief about what $\theta$ are
likely to be. For example, in the linear regression case, the instrument used
to measure the outcomes may have a well-known margin of error, or the sign of
the slope may be obvious from previous experiments. The Bayesian approach to
model fitting makes use of this information by codifying the investigator's
beliefs as a \defn{prior distribution} on the parameters, denoted
$\Pr(\theta)$. Instead of considering only the likelihood, Bayesian inference
focuses on the product of the likelihood and the prior, $\Pr(\vec{y} \mid
\theta) \Pr(\theta)$. Bayes' theorem tells us that this product is related to
the \textit{posterior distribution} on $\theta$,
\[
  \Pr(\theta \mid \vec{y}) = \frac{\Pr(\vec{y} \mid \theta) \Pr(\theta)}
                                  {\Pr(\vec{y})}.
\]

@@

A popular method for circumventing the normalizing constant is the use of
\gls{MCMC} to obtain a sample from the posterior distribution. \Gls{MCMC} works
by defining a Markov chain whose states are indexed by possible model
parameters. The transition probability from state $\theta_1$ to state
$\theta_2$ is taken to be
\[
  \max\left(1, \frac{\Pr(\vec{y} \mid \theta_2) \Pr(\theta_2) q(\theta_2, \theta_1)}
                    {\Pr(\vec{y} \mid \theta_1) \Pr(\theta_2) q(\theta_1, \theta_2)} \right),
\]
where $q(\theta, \theta')$ is a symmetric \defn{proposal distribution} used in
the algorithm to generate the chain. The stationary distribution of this Markov
chain is equal to the posterior distribution on $\theta$. Therefore, if a long
enough random walk is performed on the chain, the distribution of states
visited will be a Monte Carlo approximation of $\Pr(\theta \mid \vec{y})$, from
which we can calculate statistics of interest. Actually performing this random
walk is straightforward and can be accomplished via the Metropolis-Hastings
algorithm (\cref{alg:mh}).

\begin{algorithm}
  \caption{Metropolis-Hastings algorithm for Markov chain Monte Carlo.}
  \begin{algorithmic}
    \State Draw $\theta$ according to $\Pr(\theta)$
    \Loop
      \State Propose $\theta'$ according to $q(\theta, \theta')$
      \State Accept $\theta \gets \theta'$ with probability
      $\max \left( 1, 
       \dfrac{\Pr(\vec{y} \mid \theta') \Pr(\theta') q(\theta', \theta)}
             {\Pr(\vec{y} \mid \theta\phantom{'}) \Pr(\theta\phantom{'}) q(\theta, \theta')}
       \right)$
    \EndLoop
  \end{algorithmic}
  \label{alg:mh}
\end{algorithm}

\subsection{Overview of ABC}
\label{subsec:abcoverview}

Most mathematical models are amenable to fitting via one or both of the
approaches, \gls{ML} or Bayesian inference, discussed above. However, there are
some, particularly in the domain of population genetics, for which calculation
of either the likelihood or the product of the likelihood and the prior may be
infeasible. For example, one or both of these quantities may be expressible
only as an intractable integral. \Gls{ABC} is designed for such cases, where
standard likelihood-based techniques for model fitting cannot be applied. Such
models are particularly prevalent in population
genetics~\autocite{beaumont2002approximate, beaumont2010approximate}.

Ordinarily, Bayesian inference targets the posterior distribution $\Pr(\theta
\mid \vec{y})$. That is, in the Bayesian framework, model parameters with
higher posterior density are ``better'' in the sense that they offer a more
credible explanation for the observed data. Approximate Bayesian computation
offers an alternative metric for parameter credibility, namely the similarity
simulated datasets to the observed data. If datasets simulated under the model
closely resemble the real data, it follows that the model is a reasonable
approximation to the real-world process generating the observed data. More
formally, suppose we have a distance measure $\rho$ defined on the space of all
possible data our model could generate. \gls{ABC} aims to sample from the joint
posterior distribution of model parameters and simulated datasets $\vec{z}$
which are within some small distance $\varepsilon$ of the observed data
$\vec{y}$,
\[
  \Pr(\theta, \vec{z} \mid \vec{y}, \varepsilon) =
  \frac{\Pr(\theta) \Pr(\vec{z} \mid \theta) \I_{A_{\varepsilon, \vec{y}}} (\vec{z})}
  {\int_{A_{\varepsilon, \vec{y}} \times \Theta} \Pr(\theta) \Pr(\vec{z} \mid \theta) \d \theta}.
\]
Here, $A_{\varepsilon, \vec{y}}$ is an $\varepsilon$-ball around $\vec{y}$ with
respect to $\rho$, $\Theta$ is the space of all possible model parameters, and
$\I$ is the indicator function~\autocite{marin2012approximate}. As we shall
see in the next section, this distribution can be sampled from exactly. The
word ``approximate'' derives from the assumption that, for a suitably chosen
distance $\rho$ and a small enough $\varepsilon$, the marginal in $\vec{z}$ of
this distribution approximates the posterior of
interest~\autocite{marin2012approximate}. That is,
\[
  \int \Pr(\theta, \vec{z} \mid \vec{y}, \varepsilon) \d\vec{z}
  \approx \Pr(\theta \mid \vec{y}).
\]
To emphasize that the distribution on the left only approximates the posterior,
we will refer to it as the \textit{\gls{ABC} target distribution}. Note that in
many formulations, the distance function $\rho$ is defined as $\rho(S(\cdot),
S(\cdot))$ where $S$ is a function which maps data points into a vector of
summary statistics. This can be useful if the data are high-dimensional or of a
complex type, but it is not strictly necessary. For instance, if the data are
numeric and of low dimension, the distance function may simply be the Euclidian
distance~\autocite{sisson2007sequential}. For more complex data,
\textcite{nakagome2013kernel} proposed the use of a kernel function (defined in
\cref{subsec:treeshape}), an approach they dubbed \defn{kernel-\gls{ABC}}.

\subsection{Algorithms for ABC}
\label{subsec:abcalg}

Algorithms for performing \gls{ABC} fall into one of three categories:
rejection, \gls{MCMC}, and \gls{SMC}. Rejection ABC is the simplest method, and
also the one which was first proposed~\autocite{rubin1984bayesianly,
tavare1997inferring}. The algorithm, outlined in \cref{alg:abcrej}, repeats the
following steps until a desired number of samples from the target distribution
are obtained. Parameter values $\theta$ are sampled according to the prior
distribution $\Pr(\theta)$. Then, a simulated dataset $\vec{z}$ is generated
from the model with the sampled parameter values. By definition, the
probability density of obtaining the particular dataset $\vec{z}$ is
$\Pr(\vec{z} \mid \theta)$. Finally, the parameters are sampled if the distance
of $\vec{z}$ from the observed data $\vec{y}$ is less than $\varepsilon$ - that
is, with probability $\I_{A_{\varepsilon, \vec{y}}}(\vec{z})$. Putting this all
together, the parameters $\theta$ are sampled with probability proportional to
\[
  \Pr(\theta) \Pr(\vec{z} \mid \theta) \I_{A_{\varepsilon, \vec{y}}}(\vec{z}),
\]
which is exactly the numerator of the \gls{ABC} target distribution. Thus,
$\theta$ represents an unbiased sample from the approximate posterior.

\begin{algorithm}
  \caption{Rejection \gls{ABC}.}
  \begin{algorithmic}
    \Loop
      \State Draw $\theta$ according to $\Pr(\theta)$
      \State Simulate a dataset $\vec{z}$ from the model with parameters $\theta$
      \If{$\rho(\vec{y}, \vec{z}) < \varepsilon$}
        \State Sample $\theta$
      \EndIf
    \EndLoop
  \end{algorithmic}
  \label{alg:abcrej}
\end{algorithm}

Rejection \gls{ABC} is easy to understand and implement, but it is not
generally computationally feasible. If the posterior is very different from the
prior, a very large number of samples may need to be taken in order to find a
simulated dataset which is close to $\vec{z}$. The inefficiency is compounded
by the curse of dimensionality - the measure of the $\varepsilon$-ball around
$\vec{y}$ decreases exponentially with the number of dimensions.
\gls{ABC}-\gls{MCMC} (\cref{alg:abcmcmc}) was designed to overcome these
hurdles~\autocite{marjoram2003markov}. The approach is similar to ordinary
Bayesian \gls{MCMC} (\cref{subsec:mfit}), except that a distance cutoff
replaces the likelihood ratio. That is, the transition probability between
states $\theta_1$ and $\theta_2$, with simulated datasets $\vec{z_1}$ and
$\vec{z_2}$, is defined as
\[
  \max\left(1, \frac{\Pr(\vec{z_2} \mid \theta_2) q(\theta_2, \theta_1)}
                    {\Pr(\vec{z_1} \mid \theta_1) q(\theta_1, \theta_2)} 
    \cdot \I_{A_{\varepsilon, \vec{y}}}(\vec{z_2}) \right).
\]

\begin{algorithm}
  \caption{ABC-MCMC.}
  \begin{algorithmic}
    \State Draw $\theta$ according to $\Pr(\theta)$
    \Loop
      \State Propose $\theta'$ according to $q(\theta, \theta')$
      \State Simulate a dataset $\vec{z'}$ according to the model with
             parameters $\theta$
      \State Accept $\theta \gets \theta'$ with probability
      $\max \left( 1, 
       \dfrac{\Pr(\theta') q(\theta', \theta)}
             {\Pr(\theta\phantom{'}) q(\theta, \theta')} 
       \cdot \I_{A_{\varepsilon, \vec{y}}}(\vec{z'}) \right)$
    \EndLoop
  \end{algorithmic}
  \label{alg:abcmcmc}
\end{algorithm}
